Meta analysis of differential expression results from AMP-AD (Control vs AD, reprocessed RNASeq)

### 2017-05-10 code/Rmd/MetaAnalysis.Rmd
Temp commit

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8456631',
                   MSBB = 'syn8484996',
                   MAYO = 'syn8466814')
covariates = lapply(covariates.ids, downloadFile)

covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
                Study = 'ROSMAP',
                BrainRegion = 'DLPFC') %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MSBB = covariates$MSBB %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::filter(Diagnosis %in% c('SD', 'ND')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Diagnosis, levels = c('ND','SD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MAYO',
                Dx = factor(Dx, levels = c('Control','AD')),
                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)
```

### Get differential expression data from synapse
```{r diffexp}
diff.exp.ids = c(ROSMAP = 'syn8456721',
                 MSBB = 'syn8502894',
                 MAYO = 'syn8468023')
diff.exp = lapply(diff.exp.ids, downloadFile)

diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')

tmp = diff.exp %>% 
  rbindlist(use.names = T, fill = T) 
tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2 are')
tmp %>%
  group_by(Study, Region, Direction, Comparison) %>%
  summarise(count = n()) %>%
  spread(Direction, count) %>%
  kable

pl = list()
for(region in c('ROSMAP', 'MSSM', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

### Get cell type fraction from synapse
```{r celltype.frac}
celltype.fraction.ids = c(ROSMAP = 'syn8124520',
                          MSBB = 'syn8146376',
                          MAYO = 'syn8135250')
celltype.fraction = lapply(celltype.fraction.ids, downloadFile)

@@

### Pathway enrichment analysis
```{r pathway.enrich}
setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
  filter(adj.P.Val <= 0.05, Direction != 'NONE') %>%
  dlply(.(Region, Direction), .fun = function(x){
    setdiff(unique(x$hgnc_symbol),c(''))
  })

tmp1 = rankprod.pfp$pfp[intersect(intersect(tmp$fb.pval, tmp$min.pval), tmp$rankprod.pval),] <= 0.05
setsToTest$META.UP = rownames(tmp1)[which(tmp1[,1])]
setsToTest$META.DOWN = rownames(tmp1)[which(tmp1[,2])]

# Convert ensembl id to hgnc_symbols
setsToTest = lapply(setsToTest, function(x, Ensemble2HGNC){
  Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$ensembl_gene_id %in% x] %>%
    unique()
}, Ensemble2HGNC)

# Get Background genes
backgroundGenes = unique(rbindlist(diff.exp, use.names = T, fill = T) %>% 
                           dplyr::select(hgnc_symbol) %>%
                           unlist)

############################################################################################################
# Function to perform Fishers enrichment analysis
fisherEnrichment <- function(genesInSignificantSet, # A character vector of differentially expressed or some significant genes to test
                             genesInGeneSet, # A character vector of genes in gene set like GO annotations, pathways etc...
                             genesInBackground # Background genes that are 
){
  genesInSignificantSet = intersect(genesInSignificantSet, genesInBackground) # back ground filtering
  genesInNonSignificantSet = base::setdiff(genesInBackground, genesInSignificantSet)
  genesInGeneSet = intersect(genesInGeneSet, genesInBackground) # back ground filtering
  genesOutGeneSet = base::setdiff(genesInBackground,genesInGeneSet)

  pval = fisher.test(
    matrix(c(length(intersect(genesInGeneSet, genesInSignificantSet)),             
             length(intersect(genesInGeneSet, genesInNonSignificantSet)),
             length(intersect(genesOutGeneSet, genesInSignificantSet)),
             length(intersect(genesOutGeneSet, genesInNonSignificantSet))), 
           nrow=2, ncol=2), alternative="greater")
  OR = (length(intersect(genesInGeneSet, genesInSignificantSet)) * length(intersect(genesOutGeneSet, genesInNonSignificantSet))) / (length(intersect(genesInGeneSet, genesInNonSignificantSet)) * length(intersect(genesOutGeneSet, genesInSignificantSet))) %>% as.numeric()
  return(data.frame(pval = pval$p.value,
                    ngenes = length(genesInGeneSet),
                    noverlap = length(intersect(genesInGeneSet, genesInSignificantSet)),
                    Odds.Ratio = OR,
                    Genes = paste(intersect(genesInGeneSet, genesInSignificantSet), collapse = '|')))}
############################################################################################################

############################################################################################################
#### Get gene sets ####
# Download enrichr gene sets from synapse
GL_OBJ = synGet('syn4867851')
ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
load(GL_OBJ@filePath)

gsets = c("BioCarta_2015", "GO_Biological_Process", "Reactome_2015", "TargetScan_microRNA", "WikiPathways_2015")
GeneSets.Enrichr = GeneSets[gsets]

GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
                   AD = GeneSets['AD:GeneticLoci'])
GeneSets = c(GeneSets.Enrichr, GeneSets.CM)

# Filter gene list
GeneSets = filterGeneSets(GeneSets, backgroundGenes, minSize = 3, maxSize = 5000)

enrichResults = list()
for (name in names(setsToTest)){
  genesInModule = setsToTest[[name]]  
  enrichResults[[name]] = lapply(GeneSets,
                                    function(x, genesInModule, genesInBackground, fisherEnrichment, rownameToFirstColumn){
                                      tmp = as.data.frame(t(sapply(x, fisherEnrichment, genesInModule, genesInBackground)))
                                      tmp = rownameToFirstColumn(tmp,'GeneSetName')
                                      return(tmp)
                                    }, unique(genesInModule), unique(backgroundGenes), fisherEnrichment, rownameToFirstColumn) %>%
    rbindlist(use.names=TRUE, idcol = 'Category') %>%
    dplyr::mutate(fdr = p.adjust(pval, 'fdr'))
  writeLines(paste0('Completed ',name))  
}

# Write results to file
tmp1 = rbindlist(enrichResults, use.names = TRUE, idcol = 'ComparisonName', fill = TRUE) %>%
  filter(!is.na(Category)) %>%
  filter(fdr <= 0.01)

tmp2 = dplyr::select(tmp1, ComparisonName, GeneSetName, Odds.Ratio) %>%
  tidyr::spread(ComparisonName,Odds.Ratio)
rownames(tmp2) = tmp2[,1]

tmp1$pval = unlist(tmp1$pval)
tmp1$ngenes = unlist(tmp1$ngenes)
tmp1$noverlap = unlist(tmp1$noverlap)
tmp1$Odds.Ratio = unlist(tmp1$Odds.Ratio)
tmp1$Genes = unlist(tmp1$Genes)

# Store logcpm in synapse
write.table(tmp1, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('enrich.tsv', 
           name = 'Pathway Enrichment Results (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```

>>>>

### Get curated gene list
```{r curate.genelist}
# Download enrichr gene sets from synapse
GL_OBJ = synGet('syn4867851')
ALL_USED_IDs = c(ALL_USED_IDs, GL_OBJ$properties$id)
load(GL_OBJ@filePath)

gsets = c("GO_Biological_Process", "Reactome_2015", "WikiPathways_2015")
GeneSets.Enrichr = GeneSets[gsets]

GeneSets.CM = list(CellTypeMarkers = GeneSets[grep('Zhang', names(GeneSets))],
                   AD = GeneSets['AD:GeneticLoci'])
GeneSets = c(GeneSets.Enrichr, GeneSets.CM) # 
```

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.ids = c(ROSMAP = 'syn8456631',
                   MSBB = 'syn8484996',
                   MAYO = 'syn8466814')
all.used.ids = covariates.ids
covariates = lapply(covariates.ids, downloadFile)

covariates$ROSMAP$BrainRegion = 'DLPFC'
covariates$MSBB = covariates$MSBB %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
covariates$MAYO = covariates$MAYO %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

covariates = covariates %>% 
  rbindlist(use.names = T, fill = T, idcol = 'Study') %>%
  dplyr::select(SampleID, Diagnosis, BrainRegion)

kable(table(covariates$Diagnosis, covariates$BrainRegion))
```

### Get differential expression data from synapse
```{r diffexp, fig.width=10, fig.height=15}
diff.exp.ids = c(ROSMAP = 'syn8456721',
                 MSBB = 'syn8502894',
                 MAYO = 'syn8468023')
all.used.ids = c(all.used.ids, diff.exp.ids)
diff.exp = lapply(diff.exp.ids, downloadFile)

diff.exp$ROSMAP = diff.exp$ROSMAP %>%
  setnames(c('Region'), c('BrainRegion'))

tmp = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(Comparison == 'CONTROL-AD')

tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

writeLines('Number of differentially expressed genes at an FDR of 0.05')
tmp %>%
  group_by(Study, BrainRegion, Direction, Comparison) %>%
  summarise(count = n()) %>%
  spread(Direction, count) %>%
  kable

pl = list()
for(region in c('ROSMAP', 'MSSM', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+BrainRegion) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

### Pathway enrichment analysis (individual brain regions)
```{r pathway.enrich}
# setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
#   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
#   dlply(.(BrainRegion), .fun = function(x){
#     setdiff(unique(x$hgnc_symbol),c(''))
#   })

setsToTest = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) > log2(1.2)) %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, levels = c(-1,1),
                                   labels = c('-1' = 'DOWN', '1' = 'UP'))) %>%
  dlply(.(BrainRegion, Direction), .fun = function(x){
    setdiff(unique(x$hgnc_symbol),c(''))
  })

# Filter gene list
GeneSets = mapply(function(x,y){
  names(x) = paste(y,names(x),sep = ':')
  return(x)
},GeneSets, names(GeneSets), SIMPLIFY = F) %>%
  Reduce(c,.)
GeneSets = GeneSets[setdiff(seq(1,length(GeneSets)), c(grep('Mus musculus', names(GeneSets))))]

GeneSets = lapply(GeneSets, function(x, backgroundGenes){
  x = intersect(x,backgroundGenes)
}, backgroundGenes)
GeneSets = GeneSets[sapply(GeneSets, length) >= 15 & sapply(GeneSets, length) <= 1000]

enrichResults = plyr::ldply(setsToTest, .fun = function(genesInGeneSet, geneSets, backgroundGenes){
  plyr::ldply(geneSets, .fun = fisherEnrichment, genesInGeneSet, backgroundGenes, .parallel = T) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, GeneSets, backgroundGenes, 
.parallel = F, .progress = 'text', .paropts = list(.packages = c('plyr','dplyr','CovariateAnalysis')))

tmp = dplyr::select(enrichResults, BrainRegion, .id, fdr) %>%
  dplyr::filter(fdr <= 0.05) %>%
  tidyr::spread(BrainRegion, fdr)

# Store logcpm in synapse
write.table(tmp, file = 'enrich.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('enrich.tsv', 
           name = 'Pathway Enrichment Results (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```

### Get cell type fraction from synapse
```{r celltype.frac, eval = FALSE}
celltype.fraction.ids = c(ROSMAP = 'syn8124520',
                          MSBB = 'syn8146376',
                          MAYO = 'syn8135250')
celltype.fraction = lapply(celltype.fraction.ids, downloadFile)

@@
